\emph{Gaia} DR2 identifies VFTS 682 as the most massive runaway star known to date

### 2018-05-13 VFTS682.tex
fixed bibtex entries without arxiv number
added comments from Vincent, added author

\title{Gaia identifies VFTS682 as the most massive % , dynamically
                                % ejected %% to fit in one line
  runaway star known to date}

\author{M.~Renzo$^{1}$, S.~E.~de~Mink$^{1}$, D.~J.~Lennon$^{2}$, J. Bestenlehner$^{3}$, S.~Justham$^{4,5}$, F.~R.~Schneider$^6$, R.~P.~van~der~Marel$^{7,8}$, J.~Vink$^{9}$ \SdM{et al.?} \\ \SdM{[ Alphabetical after Lennon. Possibly add further last-minute coauthors]},  }

@@

\affil{$^{1}$Astronomical Institute Anton Pannekoek, University of
    Amsterdam, 1098 XH Amsterdam, The Netherlands \\
    $^{2}$ ESA, European Space Astronomy Centre, Apdo. de Correos 78,
    E-28691 Villanueva de la Ca\~nada, Madrid, Spain \\
    $^3$ {Department of Physics and Astronomy, Hicks Building,
      Hounsfield Road, University of Sheffield, Sheffield S3 7RH, UK}\\
   $^{4}$ {School of Astronomy \& Space Science, University of the Chinese
   Academy of Sciences, Beijing 100012, China}\\
    $^5$ {National Astronomical Observatories, Chinese Academy of
      Sciences, Beijing 100012, China}\\
     $^6$ {Department of Physics, University of Oxford, Keble Road,
       Oxford OX1 3RH, UK} \\
     $^{7}$ {Space Telescope Science Institute, 3700 San Martin Drive,
       Baltimore, MD 21218, USA}\\
     $^{8}$ {Center for Astrophysical Sciences, Department of Physics \& Astronomy, Johns Hopkins University, Baltimore, MD 21218, USA}\\
     $^9$ {Armagh Observatory, College Hill, Armagh BT61 9DG, UK}
}

%\thanks{Corresponding author:  M.~Renzo, \href{mailto:m.renzo@uva.nl}{m.renzo@uva.nl}}
\date{}
\begin{abstract}
The formation and fates of the most massive stars remains one of the most elusive questions in stellar astrophysics. 
%
Previous spectroscopic studies identified VFTS682 as one of the most massive stars known. It resides in the field of the 30 Dor region in the Large Magellanic Cloud at a projected distance of 29\,pc of the star cluster R136. The lack of a detected offset in radial velocity derived from the spectra, led to speculations whether such extreme stars can form in relative isolation. 
%
We use the unprecedented proper motions from the second Gaia data
release (DR2) to investigate the kinematics of this object. We show
that VFTS 682 is a bonafide runaway star moving away from the star
cluster R136 with a relative
projected two-dimensional speed of about $\sim$$34\pm1 \kms$,
implying a three-dimensional speed of about $\sim$$40\pm20\kms$ when
accounting for the velocity component along the line of sight.  The
inferred age, distance, and the magnitude and direction of the
velocity are consistent with a dynamical ejection scenario from the cluster. 
%
This makes VFTS 682 the most massive runaway star known to date.  It proves that star clusters are capable of ejecting their most massive members. We discuss the implications for the existence of stars with masses well above $100\Msun$ inside the cluster, the internal dynamics in dense star clusters and the final fate of this object.

@@

\keywords{stars: kinematics, stars: runaways, stars: individual: VFTS682}
\maketitle{}

@@

An interesting contribution to the debate on whether or not massive
stars form in relative isolation was presented by
\cite{bestenlehner:11} and \cite{bressert:12}, who discuss the case of the very massive star
VFTS682.   This star is located in the field of the 30 Doradus region
in the Large Magellanic Cloud (LMC) and was studied as part of the
multi-epoch spectroscopic VLT-FLAMES Tarantula Survey (VFTS) of
massive stars \citep{evans:11}.  Its spectral type is WNh5 and its 
present-day mass is $137.8^{+27.5}_{-15.9}\,M_\odot$,
\citep{schneider:18}. These correspond to an inferred initial mass of
$150.0^{+28.7}_{-17.4}\,M_\odot$ and a wind mass loss rate of
$\sim10^{-4.1\pm0.2}\,M_\odot \ \mathrm{yr}^{-1}$ inferred from the
spectral analysis not accounting for clumping
\citep[][]{bestenlehner:11}, which make it one of the
most extreme objects in the region.  It is reminiscent of the very
massive stars % , a.k.a. ``monster stars'',
identified by
\citet{crowther:10, crowther:16} in the core of the
R136 cluster.

Most remarkable is the location of VFTS682, at a projected distance of
$\sim$$29$\,pc from the core of the star cluster
R136. \citet{bestenlehner:11} discuss two possibilities for the
origin of VFTS 682. Quoting directly from their abstract: "(i) the star either formed in situ, which would have profound implications for the formation mechanism of massive stars, or (ii) VFTS 682 is a slow runaway star that originated from the dense cluster R136, which would make it the most massive runaway known to date."

% don't know where to go from here without giving away the main
% result...Stephen wants to add a citation to fujii and banerjee in
% the intro, which I also think would be fair...
% The case of VFTS682 has also spawned significant theoretical efforts
% to explain its isolation via dynamical ejection from R136, e.g.,
% \citet{fujii:11} and \citet{banerjee:12}. Both these studies suggested

Recently, \citet{platais:15,platais:18} used multi-epoch Hubble Space
Telescope imaging to identify and shed new light on runaway stars from
the 30 Doradus region based on proper motion measurements. In this study we report on our analysis of the data from the second Gaia data release \cite[DR2,][]{gaia:16,brown:18}.  We combine the radial velocity measurements from the VFTS survey \citep[][]{evans:11} with the proper motion from Gaia DR2 to reconstruct the three-dimensional velocity of VFTS682, and test the hypothesis that this star was ejected from R136.

@@

\subsection{ Overview of VFTS682 from previous studies \label{data:vfts683}}

The star VFTS682  \citep{evans:11}, located at right ascension (RA)
05$^\mathrm{h}$38$^\mathrm{m}$55.510$^\mathrm{s}$  and declination
(DEC) \mbox{-69$^\mathrm{o}$04'26.72''} \citep[][]{evans:11}
was originally classified as a young stellar object \citep{gruendl:09}
based on its mid-infrared excess.  \citet{evans:11} classified the
object as Wolf-Rayet star of spectral type WNh5 using spectra from the
VFTS survey.  The spectra available covered $\lambda$4000--7000 and
were taken at multiple epochs. Using the same dataset,
\citet{bestenlehner:11} excluded the presence of a close companion
with high confidence from the absence of radial velocity variations.

\citet{bestenlehner:11} further analyzed the spectra using the non-LTE
model atmosphere code CMFGEN \citep{hillier:98} to derive the stellar
parameters and surface abundances. They inferred an extinction
of $R_V\simeq4.7$, leading to the high luminosity
$\log_{10}(L/L_\odot) =  6.5\pm0.2$. \citet{schneider:18} estimated
a present-day mass for VFTS682 of $137.8^{+27.5}_ {-15.9}\Msun$, an
apparent age of $1.0^{+0.2}_{-0.2}$\,Myr and an inferred initial mass
of $150.0^{+28.7}_{-17.4}\Msun$, using the Bayesian analysis tool Bonnsai
\citep{schneider:17}, based on the evolutionary models from
\citet{brott:11} and \cite{kohler:15}. These values place VFTS682 around the
``canonical upper limit'' of $\sim$$150\Msun$ by \citet{figer:05}.

The star is very similar from the spectral point of view to R136a3, located inside the star cluster
R136 \citep{crowther:10}  for which \citet{crowther:16} report a
current mass estimate of $180^{+30}_{-30}\Msun$. The R136 cluster hosts
at least two more very massive WN5h stars, R136a1 and R136a2, whose
estimated current masses are even higher. However, VFTS682 stands
out by its isolation at a projected offset of 29\,pc from the center
of the cluster \citep{bestenlehner:11}.

@@

We obtain the peculiar radial velocity $\delta v_\mathrm{rad}$ of VFTS682 as the difference
between the average radial velocity of the 30 Doradus region
($270\pm10\kms$) minus the radial velocity measured from the NV $\lambda4944$
line for VFTS682  \citep[$300\pm10\kms$, ][]{bestenlehner:11}. We implicitly assume that the use
of a slightly different reference frame for $\delta v_\mathrm{rad}$ does not
introduce significant errors. We also note that the determination of
radial velocities for emission-lines stars is difficult because of the
presence of an optically thick, likely non-homogeneous, and variable
wind hiding the stellar surface.

\subsection{New data from Gaia DR2  \label{data:gaia}}
%% using
%% /scratch/mathieu/MMPS/SCRIPTS_dirty/my_script/binaries/binary_c/30Dor/vfts682/data/asu.tsv to find average

\begin{table}[t]
\begin{center}
    \caption{Astrometric parameters for VFTS682. }
  \begin{tabular}{l|c|c}
  \tableline
\tableline
    Parameter & Value & Source\\ 
    \tableline
    RA \hfill[degree] &  \phantom{-}84.73 $\pm$  0.03 & \multirow{4}{*}{Gaia DR2}\\
    DEC \hfill [degree] & -69.07 $\pm$  0.05  & \\
    $\mu_\mathrm{RA}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & \phantom{-0}1.84 $\pm$ 0.07 & \\
    $\mu_\mathrm{DEC}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & \phantom{-0}0.78 $\pm$ 0.08& \\
    $\delta v_\mathrm{rad}$  \hfill[$\kms$] & \phantom{-0}30 $\pm$ 20 & \cite{bestenlehner:11}\\
\tableline
  \end{tabular}
  \tablecomments{We truncate the errors on RA and DEC for the purpose
    of display. Readers are encouraged to consult the Gaia DR2 catalog
  for accurate values.}
    \end{center}
  \label{tab:vfts682}
\end{table}

The Gaia Data Release 2, which became available on 25 April 2018,
provides the five-parameter astrometric solution (positions,
parallaxes, and proper motion components) for more than a billion
sources \citep{brown:18}. VFTS682 is identified with the id 4657685637907503744 in the Gaia DR2
catalog\footnote{\url{https://vizier.u-strasbg.fr/viz-bin/VizieR-3?-source=I/345/gaia2}}. Gaia
reports a G-band magnitude of 15.65. The star has a
\texttt{visibility\_period} = 17, which counts how many observations have
been used to reconstruct its astrometric solution
\citep[][]{lindengren:18}, and the reported
\texttt{astrometric\_excess\_noise} = 0. These values suggest that the Gaia
data for VFTS682 are reliable. However, the effective temperature
reported in Gaia DR2 is one order of magnitude lower than what found by
\cite{bestenlehner:11}, and the best fit parallax of this star is
negative \citep[see, e.g.,][]{hogg:18}. We do not use the effective temperature of the star anywhere
in this study, and we attribute the unphysical value of the parallax
to the large distance to the LMC. Our main findings do not rely on the
parallax nor the effective temperature values reported in the Gaia DR2
catalog.

We retrieve for VFTS682 the position in RA and DEC
in the IGCS frame \cite[][]{brown:18}, its
proper motion components ($\mu_\mathrm{RA}$, and $\mu_\mathrm{DEC}$,
respectively). For the radial velocity of VFTS682 and of the 30 Doradus
region as a whole, we instead use the VFTS data
as quoted in \cite{bestenlehner:11}. \Tabref{tab:vfts682} lists the values adopted throughout
this work for each of these quantities.

We define a local standard frame of reference to derive the peculiar velocity
of VFTS682 with respect to the cluster R136 by selecting from the Gaia DR2 catalog following closely the approach of \cite{vandermarel:02,lennon:18}.
We select all the stars in a target of 0.05 degrees around R136
(NGC2070) fulfilling the following criteria. First, we require G-band
magnitude brighter than 17, correspondingly roughly to the
completeness level of the VFTS survey \citep[here we implicitly assume
G$\sim$V,][]{evans:11}. Then we require \texttt{visibility\_period} $\geq$ 5,
\texttt{astrometric\_excess\_noise} $< 1$, the errors on the proper
motion components to be smaller than 0.07\,$\mathrm{mas\
  yr^{-1}}$. At the distance to the LMC, $0.1\mathrm{mas\
  yr^{-1}}\simeq25\,\kms$ \citep[e.g.,][]{platais:18}. We
remove obvious foreground stars by requiring the absolute value of
the reported parallaxes smaller than $0.1\,\mathrm{mas\ yr^{-1}}$.
% At the distance to the
% LMC, $1\mathrm{mas\ yr^{-1}}\simeq250\,\kms$ \citep[e.g.,][]{lennon:18}, so the cut on the values
% of the proper motions removes stars that would have projected
% tangential velocities in excess of $\sim$500\,$\kms$, which are most
% likely to be foreground stars. We checked that the additional
% requirement of having parallaxes smaller than $2\,\mathrm{mas}$ does
% not reduces further our sample.
This selection yields a sample of 66 stars.% ,

We calculate the averaged proper motion components for the whole
region using 
\begin{equation}
  \label{eq:mean}
  \langle \mu_i\rangle = \frac{\sum_\mathrm{stars}\frac{1}{\Delta
      \mu_i^2}\mu_i}{\sum_\mathrm{stars} \frac{1}{\Delta \mu_i^2}} \ \ , \
  \ \Delta \langle \mu_i\rangle = \frac{\sqrt{N}}{\sum_\mathrm{stars}
    \frac{1}{\Delta \mu_i^2}} \ \ ,
\end{equation}
where $i = \mathrm{RA}, \mathrm{DEC}$, and $\Delta \mu_i$ is the error
on the proper motion component reported by Gaia. The sums run over
all the $N=66$ stars in our selected sample. We evaluate each proper motion
component separately. For simplicity, throughout this study, we assume the same
distance of $50$\,kpc to the star \citep[][]{pietrzynski:13}, and to
the 30 Doradus region as a whole. We do not consider the error bars on
the distance determination when converting proper motions into
physical velocities. The data retrieved, and the ipython notebook used for the analysis
presented here will be made available at \todo{git repo bitbucket?}.

\section{The kinematics of VFTS682}
\label{sec:results}

\subsection{Is it a runaway star?}
\label{sec:runaway}
We first address the question of whether VFTS682 is a typical star
from the kinematic point of view, or whether it is a runaway star with
a significantly large peculiar velocity compared to its surrounding population. The former is what should
be expected if it formed where we observe it today, in relative
isolation from other massive stars.

Using the stars selected as described in \Secref{data:gaia} (a
subset is shown in blue in \Figref{fig:main}), we find averaged proper motion components of
$\langle\mu_\mathrm{RA}\rangle = 1.7240\pm0.0002\,\mathrm{mas\ yr^{-1}}$ and
$\langle\mu_\mathrm{DEC}\rangle = 0.6986\pm0.0003\,\mathrm{mas\
  yr^{-1}}$. These values are
in good agreement with what is found by \cite{lennon:18}. However, we
emphasize their sensitivity to the choice of stars adopted to define
the reference frame. Subtracting these values from the
proper motion of VFTS682 (see \Tabref{tab:vfts682}), we obtain the
components of proper motion of the star relative to the surrounding region
$\delta\mu_\mathrm{RA} = 0.12\pm 0.07\,\mathrm{mas\ yr^{-1}}$ and $\delta\mu_\mathrm{DEC} =
0.08\pm 0.08\,\mathrm{mas\ yr^{-1}}$. The error budget is
dominated by the errors on the proper motion components of VFTS682.

@@

\begin{equation}
  \label{eq:speed_around}
  v_\mathrm{pec} = \sqrt{\left(\delta v_\mathrm{RA}\right)^2
    +\left(\delta v_\mathrm{DEC}\right)^2+\left(\delta
      v_\mathrm{rad}\right)^2} = 46 \pm 18 
  \kms \ .
\end{equation}
This value for the three-dimensional speed of VFTS682 with respect the
surrounding stars make it the most massive ``bona fide'' runaway star
known to date.

@@

The red arrow in \Figref{fig:main} shows the proper motion of VFTS682
relative to the region, and the lighter red arrows show the possible
range of directions within the uncertainties in the measured proper
motion. It is clear that the most likely origin of the star is R136,
as suggested based on numerical N-body simulations by \cite{fujii:11}
and \cite{banerjee:12}.

@@

\begin{equation}
  \label{eq:kin_age}
  \tau_\mathrm{kin} = \frac{d_\parallel}{v_\parallel} \simeq
  \frac{29\,\mathrm{pc}}{34\,\kms} \simeq 0.85\pm\,0.02\, \mathrm{Myr} \ \ ,
\end{equation}
where $d_\parallel =29$\,pc is the projected distance from VFTS682 to
the core of R136 \citep[][]{bestenlehner:11}, $v_\parallel \equiv \sqrt{\left(\delta v_\mathrm{RA}\right)^2
    +\left(\delta v_\mathrm{DEC}\right)^2} =34\pm
1\,\kms$ is the speed of the star relative to the cluster projected on the sky, obtained
using the relative proper motion components calculated in
\Secref{sec:runaway}.% , and we use the approximation $1 \kms \simeq 1
% \mathrm{pc \ Myr^{-1}}$.
As in the rest of this study, we neglect for
simplicity the error on the distance estimates.

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.48\textwidth]{./figures/main_plot_good}  
  \caption{The red solid arrow indicates the proper motion of VFTS682
    relative to the region, starting from the present day position of
    the star. The orange arrows indicate the possible
    directions of projected motion within the errors, and are extended
    backwards (dashed) to illustrate the uncertainty on the origin of the
    star. The blue stars
    are a subset of those we use to define the local rest
    frame. \todo{load color picture on background}}

@@

\section{Discussion}
\label{sec:discussion}
\todo{subsection on validation with HST data?Discussion on
  correlations between pmRA and pmDEC?Not sure where to start. HST
  data for VFTS682 dont seem to be public.}
\todo{Selma: Should add caveat that it may be a spurious detection.  That reconfirmation in following epochs is warranted. }

>>>>

\title{\emph{Gaia} DR2 identifies VFTS 682 as the most massive % , dynamically
                                % ejected %% to fit in one line
  runaway star known to date}

\author{M.~Renzo$^{1}$, S.~E.~de~Mink$^{1}$, D.~J.~Lennon$^{2}$,
  R.~P.~van~der~Marel$^{3,4}$, J. Bestenlehner$^{5}$, V.~H\'enault-Brunet$^{6}$, S.~Justham$^{7,8}$, F.~R.~Schneider$^9$, J.~Vink$^{10}$}

@@

\affil{$^{}$Astronomical Institute Anton Pannekoek, University of
    Amsterdam, 1098 XH Amsterdam, The Netherlands \\
    $^{}$ ESA, European Space Astronomy Centre, Apdo. de Correos 78,
    E-28691 Villanueva de la Ca\~nada, Madrid, Spain \\
      $^{}${Department of Physics and Astronomy, Hicks Building,
      Hounsfield Road, University of Sheffield, Sheffield S3 7RH, UK}\\
     $^{}$ {School of Astronomy \& Space Science, University of the Chinese
   Academy of Sciences, Beijing 100012, China}\\
    $^{}$ {National Astronomical Observatories, Chinese Academy of
      Sciences, Beijing 100012, China}\\
     $^{}$ {Department of Physics, University of Oxford, Keble Road,
       Oxford OX1 3RH, UK} \\
    $^{}$ {Space Telescope Science Institute, 3700 San Martin Drive,
       Baltimore, MD 21218, USA}\\
      {Center for Astrophysical Sciences, Department of Physics \& Astronomy, Johns Hopkins University, Baltimore, MD 21218, USA}\\
     $^{}$ {Armagh Observatory, College Hill, Armagh BT61 9DG, UK}
     $^{}$ {National Research Council, Herzberg Astronomy \& Astrophysics, 5071 West Saanich Road, Victoria, BC, V9E 2E7, Canada}     
}

%\thanks{Corresponding author:  M.~Renzo, \href{mailto:m.renzo@uva.nl}{m.renzo@uva.nl}}
\date{}
\begin{abstract}
The formation and fates of the most massive stars remains one of the most elusive questions in stellar astrophysics. 
%
Previous spectroscopic studies identified VFTS 682 as one of the most massive stars known. It resides in the field of the 30 Dor region in the Large Magellanic Cloud at a projected distance of 29\,pc of the star cluster R136. The lack of a detected offset in radial velocity derived from the spectra, led to speculations whether such extreme stars can form in relative isolation. 
%
We use the unprecedented proper motions from the second \emph{Gaia} data
release (DR2) to investigate the kinematics of this object. We show
that VFTS 682 is a bonafide runaway star moving away from the star
cluster R136 with a relative
projected two-dimensional speed of about $\sim$$34\pm1 \kms$,
implying a three-dimensional speed of about $\sim$$40\pm20\kms$ when
accounting for the velocity component along the line of sight. The
inferred age, distance, and the magnitude and direction of the
velocity are consistent with a dynamical ejection scenario from the cluster. 
%
This makes VFTS 682 the most massive runaway star known to date. It proves that star clusters are capable of ejecting their most massive members. We discuss the implications for the existence of stars with masses well above $100\Msun$ inside the cluster, the internal dynamics in dense star clusters and the final fate of this object.

@@

\keywords{stars: kinematics, stars: runaways, stars: individual: VFTS 682}
\maketitle{}

@@

An interesting contribution to the debate on whether or not massive
stars form in relative isolation was presented by
\cite{bestenlehner:11} and \cite{bressert:12}, who discuss the case of the very massive star
VFTS 682.  This star is located in the field of the 30 Doradus region
in the Large Magellanic Cloud (LMC) and was studied as part of the
multi-epoch spectroscopic VLT-FLAMES Tarantula Survey (VFTS) of
massive stars \citep{evans:11}. Its spectral type is WNh5 and its 
present-day mass is $137.8^{+27.5}_{-15.9}\,M_\odot$,
\citep{schneider:18}. These correspond to an inferred initial mass of
$150.0^{+28.7}_{-17.4}\,M_\odot$ and a wind mass loss rate of
$\sim10^{-4.1\pm0.2}\,M_\odot \ \mathrm{yr}^{-1}$ inferred from the
spectral analysis not accounting for clumping
\citep[][]{bestenlehner:11}, which make it one of the
most extreme objects in the region. It is reminiscent of the very
massive stars % , a.k.a. ``monster stars'',
identified by
\citet{crowther:10, crowther:16} in the core of the
R136 cluster.

Most remarkable is the location of VFTS 682, at a projected distance of
$\sim$$29$\,pc from the core of the star cluster
R136. \citet{bestenlehner:11} discuss two possibilities for the
origin of VFTS 682. Quoting directly from their abstract: "(i) the star either formed in situ, which would have profound implications for the formation mechanism of massive stars, or (ii) VFTS 682 is a slow runaway star that originated from the dense cluster R136, which would make it the most massive runaway known to date."

% don't know where to go from here without giving away the main
% result...Stephen wants to add a citation to fujii and banerjee in
% the intro, which I also think would be fair...
% The case of VFTS 682 has also spawned significant theoretical efforts
% to explain its isolation via dynamical ejection from R136, e.g.,
% \citet{fujii:11} and \citet{banerjee:12}. Both these studies suggested

Recently, \citet{platais:15,platais:18} used multi-epoch Hubble Space
Telescope imaging to identify and shed new light on runaway stars from
the 30 Doradus region based on proper motion measurements. In this study we report on our analysis of the data from the second \emph{Gaia} data release \cite[DR2,][]{gaia:16,brown:18}. We combine the radial velocity measurements from the VFTS survey \citep[][]{evans:11} with the proper motion from \emph{Gaia} DR2 to reconstruct the three-dimensional velocity of VFTS 682, and test the hypothesis that this star was ejected from R136.

@@

\subsection{ Overview of VFTS 682 from previous studies \label{data:vfts683}}

The star VFTS 682  \citep{evans:11}, located at right ascension (RA)
05$^\mathrm{h}$38$^\mathrm{m}$55.510$^\mathrm{s}$  and declination
(DEC) \mbox{-69$^\mathrm{o}$04'26.72''} \citep[][]{evans:11}
was originally classified as a young stellar object \citep{gruendl:09}
based on its mid-infrared excess. \citet{evans:11} classified the
object as Wolf-Rayet star of spectral type WNh5 using spectra from the
VFTS survey. The spectra available covered $\lambda$4000--7000 and
were taken at multiple epochs. Using the same dataset,
\citet{bestenlehner:11} excluded the presence of a close companion
with high confidence from the absence of radial velocity variations.

\citet{bestenlehner:11} further analyzed the spectra using the non-LTE
model atmosphere code CMFGEN \citep{hillier:98} to derive the stellar
parameters and surface abundances. They inferred an extinction
of $R_V\simeq4.7$, leading to the high luminosity
$\log_{10}(L/L_\odot) =  6.5\pm0.2$. \citet{schneider:18} estimated
a present-day mass for VFTS 682 of $137.8^{+27.5}_ {-15.9}\Msun$, an
apparent age of $1.0^{+0.2}_{-0.2}$\,Myr and an inferred initial mass
of $150.0^{+28.7}_{-17.4}\Msun$, using the Bayesian analysis tool Bonnsai
\citep{schneider:17}, based on the evolutionary models from
\citet{brott:11} and \cite{kohler:15}. These values place VFTS 682 around the
``canonical upper limit'' of $\sim$$150\Msun$ by \citet{figer:05}.

The star is very similar from the spectral point of view to R136a3, located inside the star cluster
R136 \citep{crowther:10}  for which \citet{crowther:16} report a
current mass estimate of $180^{+30}_{-30}\Msun$. The R136 cluster hosts
at least two more very massive WN5h stars, R136a1 and R136a2, whose
estimated current masses are even higher. However, VFTS 682 stands
out by its isolation at a projected offset of 29\,pc from the center
of the cluster \citep{bestenlehner:11}.

@@

We obtain the peculiar radial velocity $\delta v_\mathrm{rad}$ of VFTS 682 as the difference
between the average radial velocity of the 30 Doradus region
($270\pm10\kms$) minus the radial velocity measured from the NV $\lambda4944$
line for VFTS 682  \citep[$300\pm10\kms$, ][]{bestenlehner:11}. We implicitly assume that the use
of a slightly different reference frame for $\delta v_\mathrm{rad}$ does not
introduce significant errors. We also note that the determination of
radial velocities for emission-lines stars is difficult because of the
presence of an optically thick, likely non-homogeneous, and variable
wind hiding the stellar surface.

\subsection{New data from \emph{Gaia} DR2  \label{data:gaia}}
%% using
%% /scratch/mathieu/MMPS/SCRIPTS_dirty/my_script/binaries/binary_c/30Dor/vfts682/data/asu.tsv to find average

\begin{table}[t]
\begin{center}
    \caption{Astrometric parameters for VFTS 682. }
  \begin{tabular}{l|c|c}
  \tableline
\tableline
    Parameter & Value & Source\\ 
    \tableline
    RA \hfill[degree] &  \phantom{-}84.73 $\pm$  0.03 & \multirow{4}{*}{\emph{Gaia} DR2}\\
    DEC \hfill [degree] & -69.07 $\pm$  0.05  & \\
    $\mu_\mathrm{RA}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & \phantom{-0}1.84 $\pm$ 0.07 & \\
    $\mu_\mathrm{DEC}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & \phantom{-0}0.78 $\pm$ 0.08& \\
    $\delta v_\mathrm{rad}$  \hfill[$\kms$] & \phantom{-0}30 $\pm$ 20 & \cite{bestenlehner:11}\\
\tableline
  \end{tabular}
  \tablecomments{We truncate the decimals on RA and DEC and the
    relative errors for the purpose
    of display. Readers are encouraged to consult the \emph{Gaia} DR2 catalog
  for accurate values.}
    \end{center}
  \label{tab:vfts682}
\end{table}

The \emph{Gaia} Data Release 2, which became available on 25 April 2018,
provides the five-parameter astrometric solution (positions,
parallaxes, and proper motion components) for more than a billion
sources \citep{brown:18}. VFTS 682 is identified with the source id 4657685637907503744 in the \emph{Gaia} DR2
catalog\footnote{\url{https://vizier.u-strasbg.fr/viz-bin/VizieR-3?-source=I/345/gaia2}}. \emph{Gaia}
reports a G-band magnitude of 15.65. The star has a
\texttt{visibility\_period} = 17, which counts how many observations have
been used to reconstruct its astrometric solution
\citep[][]{lindengren:18}, and the reported
\texttt{astrometric\_excess\_noise} = 0. These values suggest that the \emph{Gaia}
data for VFTS 682 are reliable. However, the effective temperature
reported in \emph{Gaia} DR2 is one order of magnitude lower than what found by
\cite{bestenlehner:11}, and the best fit parallax of this star is
negative \citep[see, e.g.,][]{hogg:18}. We do not use the effective temperature of the star anywhere
in this study, and we attribute the unphysical value of the parallax
to the large distance to the LMC. Our main findings do not rely on the
parallax nor the effective temperature values reported in the \emph{Gaia} DR2
catalog.

We retrieve for VFTS 682 the position in RA and DEC
in the IGCS frame \cite[][]{brown:18}, its
proper motion components ($\mu_\mathrm{RA}$, and $\mu_\mathrm{DEC}$,
respectively). For the radial velocity of VFTS 682 and of the 30 Doradus
region as a whole, we instead use the VFTS data
as quoted in \cite{bestenlehner:11}. \Tabref{tab:vfts682} lists the values adopted throughout
this work for each of these quantities.

We define a local standard frame of reference to derive the peculiar velocity
of VFTS 682 with respect to the cluster R136 by selecting from the \emph{Gaia} DR2 catalog following closely the approach of \cite{vandermarel:02,lennon:18}.
We select all the stars in a target of 0.05 degrees around R136
(NGC2070) fulfilling the following criteria. First, we require G-band
magnitude brighter than 17, correspondingly roughly to the
completeness level of the VFTS survey \citep[here we implicitly assume
G$\sim$V,][]{evans:11}. Then we require \texttt{visibility\_period} $\geq$ 5,
\texttt{astrometric\_excess\_noise} $< 1$, the errors on the proper
motion components to be smaller than 0.07\,$\mathrm{mas\
  yr^{-1}}$. At the distance to the LMC, $0.1\mathrm{mas\
  yr^{-1}}\simeq25\,\kms$ \citep[e.g.,][]{platais:18}. We
remove obvious foreground stars by requiring the absolute value of
the reported parallaxes smaller than $0.1\,\mathrm{mas\ yr^{-1}}$.
% At the distance to the
% LMC, $1\mathrm{mas\ yr^{-1}}\simeq250\,\kms$ \citep[e.g.,][]{lennon:18}, so the cut on the values
% of the proper motions removes stars that would have projected
% tangential velocities in excess of $\sim$500\,$\kms$, which are most
% likely to be foreground stars. We checked that the additional
% requirement of having parallaxes smaller than $2\,\mathrm{mas}$ does
% not reduces further our sample.
This selection yields a sample of 66 stars.% ,

We calculate the averaged proper motion components for the whole
region using 
\begin{equation}
  \label{eq:mean}
  \langle \mu_i\rangle = \frac{\sum_\mathrm{stars}\frac{1}{\Delta
      \mu_i^2}\mu_i}{\sum_\mathrm{stars} \frac{1}{\Delta \mu_i^2}} \ \ , \
  \ \Delta \langle \mu_i\rangle = \frac{\sqrt{N}}{\sum_\mathrm{stars}
    \frac{1}{\Delta \mu_i^2}} \ \ ,
\end{equation}
where $i = \mathrm{RA}, \mathrm{DEC}$, and $\Delta \mu_i$ is the error
on the proper motion component reported by \emph{Gaia}. The sums run over
all the $N=66$ stars in our selected sample. We evaluate each proper motion
component separately. For simplicity, throughout this study, we assume the same
distance of $50$\,kpc to the star \citep[][]{pietrzynski:13}, and to
the 30 Doradus region as a whole. We do not consider the error bars on
the distance determination when converting proper motions into
physical velocities. The data retrieved, and the ipython notebook used for the analysis
presented here will be made available at \todo{git repo bitbucket?}.

\section{The kinematics of VFTS 682}
\label{sec:results}

\subsection{Is it a runaway star?}
\label{sec:runaway}
We first address the question of whether VFTS 682 is a typical star
from the kinematic point of view, or whether it is a runaway star with
a significantly large peculiar velocity compared to its surrounding population. The former is what should
be expected if it formed where we observe it today, in relative
isolation from other massive stars.

Using the stars selected as described in \Secref{data:gaia} (a
subset is shown in blue in \Figref{fig:main}), we find averaged proper motion components of
$\langle\mu_\mathrm{RA}\rangle = 1.7240\pm0.0002\,\mathrm{mas\ yr^{-1}}$ and
$\langle\mu_\mathrm{DEC}\rangle = 0.6986\pm0.0003\,\mathrm{mas\
  yr^{-1}}$. These values are
in good agreement with what is found by \cite{lennon:18}. However, we
emphasize their sensitivity to the choice of stars adopted to define
the reference frame. Subtracting these values from the
proper motion of VFTS 682 (see \Tabref{tab:vfts682}), we obtain the
components of proper motion of the star relative to the surrounding region
$\delta\mu_\mathrm{RA} = 0.12\pm 0.07\,\mathrm{mas\ yr^{-1}}$ and $\delta\mu_\mathrm{DEC} =
0.08\pm 0.08\,\mathrm{mas\ yr^{-1}}$. The error budget is
dominated by the errors on the proper motion components of VFTS 682.

@@

\begin{equation}
  \label{eq:speed_around}
  v_\mathrm{pec} = \sqrt{\left(\delta v_\mathrm{RA}\right)^2
    +\left(\delta v_\mathrm{DEC}\right)^2+\left(\delta
      v_\mathrm{rad}\right)^2} = 46 \pm 18 
  \kms \ .
\end{equation}
This value for the three-dimensional speed of VFTS 682 with respect the
surrounding stars make it the most massive ``bona fide'' runaway star
known to date.

@@

The red arrow in \Figref{fig:main} shows the proper motion of VFTS 682
relative to the region, and the lighter red arrows show the possible
range of directions within the uncertainties in the measured proper
motion. It is clear that the most likely origin of the star is R136,
as suggested based on numerical N-body simulations by \cite{fujii:11}
and \cite{banerjee:12}.

@@

\begin{equation}
  \label{eq:kin_age}
  \tau_\mathrm{kin} = \frac{d_\parallel}{v_\parallel} \simeq
  \frac{29\,\mathrm{pc}}{34\,\kms} \simeq 0.85\pm\,0.02\, \mathrm{Myr} \ \ ,
\end{equation}
where $d_\parallel =29$\,pc is the projected distance from VFTS 682 to
the core of R136 \citep[][]{bestenlehner:11}, $v_\parallel \equiv \sqrt{\left(\delta v_\mathrm{RA}\right)^2
    +\left(\delta v_\mathrm{DEC}\right)^2} =34\pm
1\,\kms$ is the speed of the star relative to the cluster projected on the sky, obtained
using the relative proper motion components calculated in
\Secref{sec:runaway}.% , and we use the approximation $1 \kms \simeq 1
% \mathrm{pc \ Myr^{-1}}$.
As in the rest of this study, we neglect for
simplicity the error on the distance estimates.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=0.48\textwidth]{./figures/main_plot_good}  
  \caption{The red solid arrow indicates the proper motion of VFTS 682
    relative to the region, starting from the present day position of
    the star. The orange arrows indicate the possible
    directions of projected motion within the errors, and are extended
    backwards (dashed) to illustrate the uncertainty on the origin of the
    star. The blue stars
    are a subset of those we use to define the local rest
    frame. \todo{load color picture on background}}

@@

\section{Discussion}
\label{sec:discussion}
\todo{subsection on validation with HST data?Discussion on
  correlations between pmRA and pmDEC?Not sure where to start. HST
  data for VFTS 682 dont seem to be public.}
\todo{Selma: Should add caveat that it may be a spurious detection. That reconfirmation in following epochs is warranted. }
